---
title: "week_3_covid"
output:
  pdf_document: default
  html_document: default
date: "2025-02-03"
---

```{r setup, include=FALSE}
# Set default options to hide warnings, messages, and excessive output
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
 
# Load libraries
suppressPackageStartupMessages({
    library(readr)
    library(stringr)
    library(tidyverse)
    library(lubridate)
    library(kableExtra)
})
```

## COVID analysis

This document describes the work done on COVID cases dataset for CU Boulder MDS 2025

### Datasources

Data comes from following repository: "<https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/>"

Files are:

-   **US confirmed cases:** "time_series_covid19_confirmed_US.csv"

-   **Global confirmed cases** "time_series_covid19_confirmed_global.csv"

-   **US deaths:** "time_series_covid19_deaths_US.csv"

-   **Global heaths**: "time_series_covid19_deaths_global.csv"

-   ** Global population**: "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
### Data loading

```{r data_source, echo=FALSE, warning=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

# File names
file_names = c("time_series_covid19_confirmed_US.csv",  
               "time_series_covid19_confirmed_global.csv", 
               "time_series_covid19_deaths_US.csv",  
               "time_series_covid19_deaths_global.csv")

# Construct URLs
urls <- str_c(url_in, file_names)

global_cases <- read_csv(urls[2])
US_cases <- read_csv(urls[1])
global_deaths <- read_csv(urls[4])
US_deaths <- read_csv(urls[3])

```

### Clean global_cases

```{r global_cases_clean, echo=FALSE, warning=FALSE}
global_cases <- global_cases %>%
     pivot_longer(
         cols = -c(`Province/State`, `Country/Region`, Lat, Long),
         names_to = "date",
         values_to = "cases"
     ) %>%
     select(-c(Lat, Long))

summary(global_cases)
```

### Clean global deaths

```{r global_deaths, echo=FALSE, warning=FALSE}

global_deaths <- global_deaths %>%
     pivot_longer(
         cols = -c(`Province/State`, `Country/Region`, Lat, Long),
         names_to = "date",
         values_to = "deaths"
     ) %>%
     select(-c(Lat, Long))

summary(global_deaths)
```
### Merge cases and deaths

```{r global_merge, echo=FALSE, warning=FALSE}

global <- global_cases %>% 
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date=mdy(date))
```
#### Filter negatives cases (cases=0)
```{r global_filter, echo=FALSE, warning=FALSE}

global <- global %>% filter(cases>0)         
summary(global)
```
### Proccess the US_cases
```{r US_cases, echo=FALSE, warning=FALSE}
US_cases <- US_cases %>% 
  pivot_longer(cols=-(UID:Combined_Key), names_to="date", values_to = "cases")%>%
  select(Admin2:cases) %>%
  mutate(date=mdy(date)) %>%
  select(-c(Lat, Long_))
```

### Proccess the US_deaths
```{r US_deaths, echo=FALSE, warning=FALSE}
US_deaths <- US_deaths %>% 
  pivot_longer(cols=-(UID:Population), names_to="date", values_to = "deaths")%>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date)) %>%
  select(-c(Lat, Long_))
```
### MERGE US CASES WITH US DEATHS
```{r US_deaths_merge, echo=FALSE, warning=FALSE}

US <- US_cases %>%
  full_join(US_deaths)
         
summary(US)
```
### ADD OUR POPULATION DATA TO THE GLOBAL DATASET
```{r read_population_data, echo=FALSE, warning=FALSE}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url)
```

#### PREPARE THE GLOBAL DATASET
```{r add_population_to_global, echo=FALSE, warning=FALSE}
global <- global %>%
 unite("Combined_key", c(Province_State, Country_Region),
 sep = ", ",
 na.rm = TRUE,
 remove = FALSE)
```

#### JOIN global dataset with population (by Province_State and Country_Region)
```{r merge_population, echo=FALSE, warning=FALSE}
global <- global %>%
  left_join(uid, by=c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```

### Show a nice table of the global database
```{r display_global, echo=FALSE}
global %>%
  head(10) %>%
  kableExtra::kable("html") %>%
  kableExtra::kable_styling("striped", full_width = FALSE)