---
title: "NYPD shooting analysis, modelization and bias study"
output:
  pdf_document: default
  html_document: default
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
# Set default options to hide warnings, messages, and excessive output
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
 
# Load libraries
suppressPackageStartupMessages({
    library(readr)
    library(stringr)
    library(tidyverse)
    library(lubridate)
    library(kableExtra)
    library(patchwork)
    library(ggthemes)  
    library(viridis)
    library(tidymodels)  # PIPELINE
    library(ranger)      # MODELIZATION
    library(vip)         # FEATURE IMPORTANCE
})
```

## NYPD shooting analysis

This document describes the work done on NYPD shooting cases dataset for CU Boulder MDS 2025

This work tries to answer the following research questions:

-   When the shootings happens, which may lead to some correlation with the time.

-   Is there any potential bias in the data

-   It is safe using machine learning on this dataset

-   Its safe to use machine learning to estimate when a possible perpetrator is guilty

### Datasources

Data comes from following repository: "<https://data.cityofnewyork.us>\>"

File is:

-   **New York shooting cases:** "<https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD>"

    **From the NYPD description:**\
    *This is a breakdown of every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year. This data is manually extracted every quarter and reviewed by the Office of Management Analysis and Planning before being posted on the NYPD website. Each record represents a shooting incident in NYC and includes information about the event, the location and time of occurrence. In addition, information related to suspect and victim demographics is also included. This data can be used by the public to explore the nature of shooting/criminal activity.* \
    <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>

```{r data_load, echo=TRUE, warning=FALSE}

url <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

nypd_shootings <- read_csv(url, show_col_types = FALSE)

# Check structure of dataset
glimpse(nypd_shootings)
```

## Data pre-process

```{r pre_proc, echo=FALSE}
### TRANSFORM THE COLUMN TYPES
### ADD WEEK DAY
### KEEP ONLY THE COLUMNS I WANT
nypd_shootings <- nypd_shootings %>%
  select(OCCUR_DATE, OCCUR_TIME, PRECINCT, PERP_SEX, PERP_AGE_GROUP, PERP_RACE, VIC_SEX, VIC_AGE_GROUP, VIC_RACE) %>%
  mutate(
    OCCUR_DATE = as.Date(OCCUR_DATE, format="%m/%d/%Y"),       # Convert date to proper format
    OCCUR_TIME = hms::as_hms(OCCUR_TIME),                      # Convert time to hms format
    HOUR = hour(OCCUR_TIME),                                   # Extract the time and ensure 24-hour format
    DAY_OF_WEEK = weekdays(OCCUR_DATE)                         # Extract the day from the week
  )

```

### DATA AGGREGATION

```{r pre_proc_data_aggregation, echo=FALSE}
### CREATES A SHOOTINGS PER DAY TABLE
shootings_per_day <- nypd_shootings %>%
  group_by(OCCUR_DATE, DAY_OF_WEEK) %>%
  summarise(shooting_count = n(), .groups = "drop")

### CREATES A SHOOTINGS BY WEEKDAY TABLE
shootings_per_weekday <- nypd_shootings %>%
  mutate(DAY_OF_WEEK = weekdays(OCCUR_DATE)) %>%  # Extract the day of the week
  group_by(DAY_OF_WEEK) %>%
  summarise(shooting_count = n(), .groups = "drop")  # Count total shootings per weekday

shootings_per_weekday$DAY_OF_WEEK <- factor(shootings_per_weekday$DAY_OF_WEEK, 
                                           levels = c("Monday", "Tuesday", "Wednesday", 
                                                      "Thursday", "Friday", "Saturday", "Sunday"))

### CREATE SHOOTING BY TIME ###
shootings_per_time <- nypd_shootings %>%
  group_by(DAY_OF_WEEK, HOUR) %>%
  summarise(shooting_count = n(), .groups = "drop")  # Count shootings per hour per weekday

# Convert weekdays to ordered factor (Monday → Sunday)
shootings_per_time$DAY_OF_WEEK <- factor(shootings_per_time$DAY_OF_WEEK, 
                                        levels = c("Monday", "Tuesday", "Wednesday", 
                                                   "Thursday", "Friday", "Saturday", "Sunday"))
```

### PLOT THE TRAJECTORY OF THE SHOOTINGS ALONG TIME

```{r plot_shootings_trajectory, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
  ggplot(shootings_per_day, aes(x = OCCUR_DATE, y = shooting_count)) +
    geom_line(color = "blue", size = 1) +
    labs(title = "Daily Shootings in NYC (Separated by Day of the Week)",
         x = "Date",
         y = "Number of Shootings") +
    theme_minimal()
  
```

### PLOT THE TRAJECTORY OF THE SHOOTINGS BY WEEKDAY BECAUSE THE TREND IS NOT CLEAR

```{r plot_shootings_trajectory_per_weekday, echo=FALSE,fig.align='center', fig.width=6, fig.height=4}
  ggplot(shootings_per_day, aes(x = OCCUR_DATE, y = shooting_count)) +
    geom_point(color = "blue", size = 1) +
    facet_wrap(~ DAY_OF_WEEK, ncol = 2) +  # Separate graphs per weekday
    labs(title = "Daily Shootings in NYC (Separated by Day of the Week)",
         x = "Date",
         y = "Number of Shootings") +
    theme_minimal()
  
```

### PLOT THE DISTRUBUTION PER WEEK DAY TO UNDERSTAND IF ITS SOME PATTERN THERE

```{r plot_weekday_distribution_trajectory, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
scatter_plot <- ggplot(shootings_per_weekday, aes(x = DAY_OF_WEEK, y = shooting_count)) +
  geom_point(color = "blue", size = 4, alpha = 1) +  # Scatter points
  labs(title = "Shootings by Day of the Week in NYC",
       x = "Day of the Week",
       y = "Number of Shootings") +
  theme_minimal()

# I'm interested in the variability per day
box_plot <- ggplot(shootings_per_day, aes(x = DAY_OF_WEEK, y = shooting_count)) +
  geom_boxplot(fill = "skyblue", alpha = 0.6) +
  labs(title = "Distribution of Daily Shootings per Weekday",
       x = "Day of the Week",
       y = "Number of Shootings per Day") +
  theme_minimal()

scatter_plot + box_plot + plot_layout(ncol = 2)
```

Its a clear pattern on the weekends, more shoots.\
**There are more shootings during the weekends**.\
Also, the boxplot confirms the scatter and my insight is: v**iolent crime increases on weekends**

Indicators:

-   Higher medians

-   More variability (more outliers and wider boxes)

-   During the weekdays, the shooting number is more stable but occurs

Observation:

Friday and Monday also shows more shootings which is consistent with the weekend proximity.

Hypothesis: Most of the shootings happens because of nightlife, to test it I'll:

-   Plot the scatter day of week vs time of the shooting being the size of the marker the shooting count.

-   This should show a clear pattern on the night during weekends and also Friday.

\

```{r plot_preccint_distribution_trajectory, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
  ggplot(shootings_per_time, aes(x = DAY_OF_WEEK, y = HOUR, size = shooting_count, color = shooting_count)) +
    geom_point(alpha = 0.9) +
    scale_size(range = c(2, 12)) +
    scale_color_viridis_c(option = "magma", direction = -1) +
    scale_y_continuous(breaks = seq(0, 23, 2)) +  # Label every 2 hours
    scale_size(range = c(2, 10)) + 
    labs(title = "Shootings by Day of the Week and Time",
         x = "Day of the Week",
         y = "Hour of the Day",
         size  = "Shooting Count",
         color = "Shooting Count",
         ) +
    theme_minimal()
```

Insight: The crimes happens moslty during weekend during the night. Its associated with nightlife.\
Future work: Check **where (using location)** the shootings happens the most to see if its close to bars.\
This can help the police to search for dangerous places.\

### POSSIBLE BIASES

As is a sensitive topic, its important to understand the potential risk of bias. For identifying them I'll:

-   Train an explainable model using all the features when the perpetrator is identified

-   Explain the values of the features that contribute the most.

This should show the potential biases if only this dataset is used for training.

#### Modelization

I'll train a Random Forest, the split will be 70% training and 30% test

```{r modelization}
### FEATURE SELECTION AND REMOVE ALL NANs
    crime_data <- nypd_shootings %>%
      select(PRECINCT, OCCUR_TIME, PERP_SEX, PERP_AGE_GROUP, PERP_RACE, VIC_SEX, VIC_AGE_GROUP, VIC_RACE) %>%
      drop_na()
    
### TRANSFORM THE CATEGORICAL VARIABLES INTO TRAINABLE ONES
  crime_data <- crime_data %>%
  mutate(across(c(PRECINCT, PERP_SEX, PERP_AGE_GROUP, PERP_RACE, VIC_SEX, VIC_AGE_GROUP, VIC_RACE), as.factor))

### SPLIT THE DATASET
  set.seed(123)
  crime_split <- initial_split(crime_data, prop = 0.70)
  crime_train <- training(crime_split)
  crime_test <- testing(crime_split)

### TRAIN THE RANDOM FOREST
  model <- rand_forest(mode = "classification") %>%
    set_engine("ranger", importance="permutation") %>%
    fit(PRECINCT ~ ., data = crime_train)

```

#### SHOW FEATURE IMPORTANCE TO IDENTIFY THE BIASES

```{r feature_importance, fig.align='center', fig.width=6, fig.height=4}
vip(model)
```

#### MODEL CORRECTION TO IDENTIFY BIASES

-   We already estabilshed that the time its important, but not useful to confirm the bias that its appearing (RACE and PERPETRATOR age group)

    ```{r bias_identification_model_correction, fig.align='center', fig.width=6, fig.height=4}

    ### FEATURE SELECTION AND REMOVE ALL NANs
        crime_data <- nypd_shootings %>%
          select(PRECINCT, PERP_SEX, PERP_AGE_GROUP, PERP_RACE, VIC_SEX, VIC_AGE_GROUP, VIC_RACE) %>%
          drop_na()
        
    ### TRANSFORM THE CATEGORICAL VARIABLES INTO TRAINABLE ONES
      crime_data <- crime_data %>%
      mutate(across(c(PRECINCT, PERP_SEX, PERP_AGE_GROUP, PERP_RACE, VIC_SEX, VIC_AGE_GROUP, VIC_RACE), as.factor))

    ### SPLIT THE DATASET
      set.seed(123)
      crime_split <- initial_split(crime_data, prop = 0.70)
      crime_train <- training(crime_split)
      crime_test <- testing(crime_split)

    ### TRAIN THE RANDOM FOREST
      corrected_model <- rand_forest(mode = "classification") %>%
        set_engine("ranger", importance="permutation") %>%
        fit(PRECINCT ~ ., data = crime_train)
    ```

#### CORRECTED FEATURE IMPORTANCE

```{r corrected_feature_importance, fig.align='center', fig.width=6, fig.height=4}
vip(corrected_model)
```

### CHECK PERPETRATOR RACE SHOOTINGS DISTRIBUTION

```{r explain_bias_by_race_distribution, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
### AGGREGATE BY PERPETRATOR RACE WHEN ITS KNOW (NOT NA) ###
shootings_by_perpetrator_race <- nypd_shootings %>%
  filter(!is.na(PERP_RACE)) %>%
  group_by(PERP_RACE) %>%
  summarise(shooting_count = n(), .groups = "drop") %>%
  arrange(desc(shooting_count))

ggplot(shootings_by_perpetrator_race, aes(x = reorder(factor(PERP_RACE), shooting_count), y = shooting_count, fill = shooting_count)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "lightblue", high = "darkred") +  # Color gradient for impact
  labs(title = "Perpetrator race int the dataset for Shootings in NYC",
       x = "Precinct",
       y = "Total Shootings") +
  theme_minimal() +
  coord_flip() 
```

#### FINALLY LETS BRIEFLY ANALYZE THE RELATIONSHIP BETWEEN PERPETRATOR RACE AND VICTIM RACE

```{r race_to_race_shootings, echo=FALSE, fig.align='center', fig.width=6, fig.height=4}
shootings_by_race <- nypd_shootings %>%
  filter(!is.na(PERP_RACE) & !is.na(VIC_RACE)) %>%  # Remove NA values
  group_by(PERP_RACE, VIC_RACE) %>%
  summarise(shooting_count = n(), .groups = "drop")

  ggplot(shootings_by_race, aes(x = PERP_RACE, y = VIC_RACE, size = shooting_count, color = shooting_count)) +
    geom_point(alpha = 0.8) +  # Add transparency
    scale_size(range = c(2, 10)) +  # Adjust marker size range
    scale_color_viridis_c() +  # Use viridis color scale for better readability
    labs(
      title = "Shootings by Perpetrator Race vs Victim Race",
      x = "Perpetrator Race",
      y = "Victim Race",
      size = "Shooting Count",
      color = "Shooting Count"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## CONCLUSIONS

-   There is an increase of shootings during the **weekends**

-   Most of the crimes happens during the **night**

-   Most of the shootings APPEARS to happen **between same race,** in this case BLACK, but **WARNING**\
    THE DATA IS **NOT** **NORMALIZED** by population, therefore, ITS AN ERRPR TO CONCLUDE THAT THE ANY RACE COMMITS MORE CRIMES - if certain race is more abundant, then the dataset is biased.

-   The dataset may be biased, therefore, the model **IS BIASED**

-   This dataset cannot be used without normalization because:

    -   Training a model with this dataset, as is, will lead to incorrect predictions, as will be enough to be black have a guilty tag.

    -   Given that most of the crimes happens between same race (according to this dataset), its highly likely that if the accused is black the victim is also black, increasing the chances to be incorrectly accused.

    -   If the perpetrator is an American Indian or Alaskan Native and guilty, your chances to get free are very high

-   Even if the model is normalized, **there is still posibility of bias by race** as, in the hypotesis that certain race is commits more crimes, the estimation will be biased to this race. Therefore, using machine learning for guiltiness estimation **is risky** and potentially racist.

## FUTURE WORK

-   Normalize the dataset by census (out of the scope of this work)

-   Study where the crimes happens
